---
title: "Romania SEM"
author: "Marissa Dyck"
date: "2024-08-08"
output:
  pdf_document:
    toc: true
  html_document:
    theme: journal
    toc: true
    toc_float: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Before you begin

## Notes

A few notes about this script.

If you are want to run through the full analysis with the published data make sure you download the whole (Romania_SEM repository)[https://github.com/marissadyck/Romania_SEM] from Marissa Dyck's GitHub. This will ensure you have all the files, data, and proper folder structure you will need to run this code and associated analyses.

Also make sure you open RStudio through the R project (Romania_SEM.Rproj) this will automatically set your working directory to the correct place (wherever you saved the repository) and ensure you don't have to change the file paths for the data. 

If you have question please email the author,   

Marissa A. Dyck   
Postdoctoral research fellow    
University of Victoria    
School of Environmental Studies     
Email: [marissadyck17@gmail.com](marissadyck17@gmail.com)    


## R and RStudio

Before starting you should ensure you have the latest version of R and RStudio downloaded. This code was generated under R version 4.2.3 and with RStudio version 2024.04.2+764.    

You can download R and RStudio [HERE](https://posit.co/download/rstudio-desktop/)   


## R markdown

This script is written in R markdown and thus uses a mix of coding markup languages and R. If you are planning to run this script with new data or make any modifications you will want to be familiar with some basics of R markdown.

Below is an R markdown cheatsheet to help you get started,    
[R markdown cheatsheet](https://www.rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf)    


## Install packages

If you don't already have the following packages installed, use the code below to install them.

```{r install packages, eval=FALSE}

install.packages('tidyverse') 
install.packages('piecewiseSEM') 
install.packages("PerformanceAnalytics")
install.packages("purrrlyr")

```


## Load libraries

Then load the packages to your library for this current session.

```{r libraries, message=FALSE}

library(piecewiseSEM) # used to run piecewise structural equation models
library(tidyverse)  # data tidying, visualization, and much more; this will load all tidyverse packages, can see complete list using tidyverse_packages()
library(PerformanceAnalytics) # generates correlation chart
library(purrrlyr) # data manipulation
```

# Data


## Import

Let's read in the data that we cleaned and formatted with the earlier code '1_RO_data_formatting'

In this code chunk I will also add columns with the scaled the predictor variables fo analysis

```{r import data}

ro_sem_dat <- read_csv('data/processed/ro_sem_dat_2018-2019.csv') %>% 
  
  # scale predictor variables
  mutate(z.s = scale(z),
         denslocalr.s = scale(denslocalr),
         tri5.s = scale(tri5),
         clc_forest.s = scale(clc_forest))

```
## Summary statistics

We will want to report some summary statistics for our data, the code chunks below will extract/calculate those details

### Total occurrences

Let's calculate the total occurrences for 8 focal species by session/season
```{r total occurrences}

ro_sem_dat %>% 
  
  # group by session 
  group_by(session) %>% 
  
  # calculate occurrences for species of interest (8)
  summarise(total_8 = sum(hare, 
                          fox, 
                          lynx, 
                          wolf, 
                          red_deer, 
                          roe_deer, 
                          wild_boar, 
                          wild_cat))

```
### Species occurrences

We will also want some summary information for each species

```{r species occurrences}

# total species occurrence summaries

ro_sem_dat %>% 
  
  # select species of interest
  select(badger:otter) %>% 
  
  # use purrr map to calculate summary stats for all columns listed
  dmap(~c(sum(.x), 
          min(.x), 
          max(.x),
          (sum(.x)/3599*100))) %>% 
  
  # create columns for new data to report
  mutate(stat = c('sum', 
                  'min', 
                  'max', 
                  'percent')) %>% 
  relocate(stat)
```

### Environmental variables

And we will want to report some summary info on the environmental variables we are using in the analysis

```{r env summary}

ro_sem_dat %>%
  
  # select columns of raw data (not scaled)
  select(z, denslocalr, tri5, clc_forest) %>% 
  
  # print summary info for each
  summary()
```

## Correlations

Before we run any models we need to test an assumption of idepended and see if any of our potential variables are correlated.

```{r correlations, warning=FALSE}

# subset data to just environmental variables for correlation tests

ro_sem_env <- ro_sem_dat %>% 
  select(tri5.s, 
         clc_forest.s, 
         denslocalr.s, 
         z.s)

# get pairwise correlation matrix

chart.Correlation(ro_sem_env,
                  histogram = TRUE,
                  pch = 19)


```

> all correlations for chosen environmental variables >0.25


# Analysis

The code chunks below define the models for our three hypotheses

* top-down  
* bottom-up   
* combined    

### Top-down

#### Hypothesis

This first chunk is our initial model as we defined it a priori based on previous research and the literature review

```{r top-down model hypothesis}

sem.top_down_a <- 
  psem(
    glm(lynx ~ denslocalr.s + clc_forest.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wolf ~ denslocalr.s + clc_forest.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(fox ~ lynx + wolf, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wild_cat ~ fox + lynx + z.s + tri5.s + clc_forest.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(hare ~ lynx + wild_cat + fox, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(red_deer ~ wolf, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(roe_deer ~ wolf + lynx, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wild_boar ~ wolf, family = poisson(link = 'log'), data = ro_sem_dat),
    (lynx %~~% wolf)
  )

summary(sem.top_down_a)
```
#### Optimized

Now, based on the tests of directed separation, we will add significant pathways to model that make biological sense and fit hypotheses (top-down) and remove non-significant pathways and pathways with coefficient estimates that do not fit the hypothesis (e.g. positive estimate for a hypothesized negative causal pathway)

```{r top-down model optimized}

sem.top_down_b <- 
  psem(
    glm(lynx ~ tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wolf ~ denslocalr.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(fox ~ lynx + wolf + z.s + clc_forest.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wild_cat ~ lynx + wolf + z.s + clc_forest.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(hare ~  wild_cat, family = poisson(link = 'log'), data = ro_sem_dat),
    #glm(red_deer ~ wolf, family = poisson(link = 'log'), data = ro_sem_dat),
    #glm(roe_deer ~ lynx, family = poisson(link = 'log'), data = ro_sem_dat),
    #glm(wild_boar ~ wolf, family = poisson(link = 'log'), data = ro_sem_dat),
    (lynx %~~% wolf),
    (wild_cat %~~% fox),
    (hare %~~% wolf),
    (hare %~~% lynx)
  )

summary(sem.top_down_b)


```


### Bottom-up

This section will define our bottom-up model

#### Hypothesis

Again, this first chunk is our initial model as we defined it a priori based on previous research and the literature review

```{r bottom-up hypothesis}

sem.bottom_up_a <- 
  psem(
    glm(wild_boar ~ clc_forest.s + denslocalr.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(roe_deer ~ clc_forest.s + denslocalr.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(red_deer ~ clc_forest.s + z.s + denslocalr.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(hare ~ clc_forest.s + denslocalr.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wild_cat ~ hare, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(fox ~ hare, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wolf ~ red_deer + wild_boar + roe_deer, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(lynx ~ hare + roe_deer, family = poisson(link = 'log'), data = ro_sem_dat),
    (roe_deer %~~% wild_boar),
    (red_deer %~~% wild_boar),
    (hare %~~% wild_boar),
    (roe_deer %~~% red_deer),
    (roe_deer %~~% hare),
    (red_deer %~~% hare)
  )

summary(sem.bottom_up_a)
```

#### Optimized

And again,  based on the tests of directed separation, we will add significant pathways to model that make biological sense and fit hypotheses (top-down) and remove non-significant pathways and pathways with coefficient estimates that do not fit the hypothesis (e.g. positive estimate for a hypothesized negative causal pathway)

```{r bottom-up optimized}

sem.bottom_up_b <- 
  psem(
    glm(wild_boar ~ clc_forest.s + denslocalr.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(roe_deer ~ clc_forest.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(red_deer ~ clc_forest.s + z.s + denslocalr.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(hare ~ tri5.s + z.s, family = poisson(link = 'log'), data = ro_sem_dat),
    #glm(wild_cat ~ wild_boar, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(fox ~ roe_deer, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wolf ~ red_deer + wild_boar, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(lynx ~ hare, family = poisson(link = 'log'), data = ro_sem_dat),
    (roe_deer %~~% wild_boar),
    (red_deer %~~% wild_boar),
    (hare %~~% wild_boar),
    (roe_deer %~~% red_deer),
    (roe_deer %~~% hare),
    (red_deer %~~% hare),
    (lynx %~~% wolf),
    (wolf %~~% fox), 
    (lynx %~~% fox),
    (wild_boar %~~% fox),
    (lynx %~~% wild_boar),
    (lynx %~~% red_deer)
  )

summary(sem.bottom_up_b)
```

### Combined

Lastly we will define a combined model that includes both bottom-up and top-down processes/interactions. This model will be based on the optimized models for both top-down and bottom-up

```{r combined model}

sem.combined <- 
  psem(
    glm(lynx ~ tri5.s + hare, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wolf ~ z.s + tri5.s + denslocalr.s + wild_boar, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(fox ~ z.s + tri5.s + wolf + lynx + roe_deer, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wild_cat ~ clc_forest.s + z.s + lynx, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(hare ~ tri5.s + z.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(red_deer ~ clc_forest.s + denslocalr.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(roe_deer ~ clc_forest.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    glm(wild_boar ~ clc_forest.s + denslocalr.s + z.s + tri5.s, family = poisson(link = 'log'), data = ro_sem_dat),
    (lynx %~~% wolf),
    (wild_boar %~~% hare),
    (roe_deer %~~% hare),
    (red_deer %~~% hare),
    (wild_boar %~~% red_deer),
    (red_deer %~~% roe_deer),
    (wild_boar %~~% roe_deer),
    (wild_cat %~~% fox),
    (red_deer %~~% lynx),
    (wolf %~~% hare),
    (wild_cat %~~% wild_boar),
    (wild_cat %~~% roe_deer),
    (wild_cat %~~% red_deer),
    (lynx %~~% wild_boar)
  )

summary(sem.combined,
        conserve = TRUE)

```

> Based on the Chi-squared and fisher's C the combined model fits the data best
